---
title: "Week Six"
output: ioslides_presentation
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Install these packages first
knitr::opts_chunk$set(message = FALSE)
library(tidyverse)
library(readr)
library(knitr)
```


# Hypothesis tests

## One sample proportion test {.build}

I flip a coin 7 times and get 6 heads. Is this convincing evidence that the coin
is biased?

$$
H_0: p = .5 \quad vs. \quad H_A: p \ne .5
$$

Let $X$ be the number of heads in 6 flips of a fair coin. $X \sim \textrm{Binom}(6, .5)$.

The p-value is $P(X \ge 6) = P(X = 6) + P(X = 7)$. (times two)

```{r}
(dbinom(6, 7, .5) + dbinom(7, 7, .5)) * 2
(1 - pbinom(5, 7, .5)) * 2
```


## Two sample t-test {.build}

Let $X \sim N(\mu_1, \sigma^2_1)$ and $Y \sim N(\mu_2, \sigma^2_2)$ and take two independent samples of size $n_1$ and $n_2$ from each. To test:

$$
H_0: \mu_1 = \mu_2 \quad vs. \quad H_A: \mu_1 \ne \mu_2
$$

form the test statistic:

$$
T = \frac{\bar{X} - \bar{Y}}{\sqrt{S^2_1/n_1 + S^2_2 / n_2}}
$$

The p-value is found by evaluating the observed $t$ in $t_{df}$.

```{r eval = FALSE}
pt(obs_t, df = n - 1) * 2
```


## Two sample z-test {.build}

Let $X \sim \textrm{Binom}(n_1, p_1)$ and $Y \sim \textrm{Binom}(n_2, p_2)$. To test:

$$
H_0: p_1 = p_2 \quad vs. \quad H_A: p_1 \ne p_2
$$

let $\hat{p}_p = \frac{X + Y}{n_1 + n_2}$ and form the test statistic:

$$
Z = \frac{\hat{p}_1 - \hat{p}_2}{\sqrt{(\hat{p}_p(1 - \hat{p}_p)/(1/n_1 + 1/n_2))}}
$$

The p-value is found by evaluating the observed $z$ in $Z$.

```{r eval = FALSE}
pnorm(obs_z) * 2
```


