---
title: "Week Four"
output: ioslides_presentation
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Install these packages first
knitr::opts_chunk$set(message = FALSE)
library(tidyverse)
library(readr)
library(knitr)
```

## Simulate Normal Data

We simulate a random sample of 1000 Normal$(\mu=0, \sigma=5)$ IID random variables.

```{r}
n <- 1000
x <- rnorm(n, mean = 0, sd = 5)
```



## Log-Likelihood Function

The log-likelihood function can be computed with the following function:

```{r}
l <- function(x, mu, sigma){
  sum(dnorm(x, mean = mu, sd = sigma, log = TRUE))
}
```

We plot the log-likelihood function for $(\mu, \sigma)$ given $\vec{x}$, saving the values in `l_surface` (code hidden).

```{r, echo=FALSE}
# Define 2D domain
mu <- seq(-1.5, 1.5, length = 500)
sigma <- seq(4.3, 5.5, length = 500)

# Compute log-likelihood
l_surface <- matrix(0, nrow = length(mu), ncol = length(sigma))
for(i in 1:nrow(l_surface)) {
  for(j in 1:ncol(l_surface)) {
    l_surface[i, j] <- l(x, mu = mu[i], sigma = sigma[j])
  }
}
```

## 

```{r}
library(plotly)
plot_ly(z = ~l_surface) %>% 
  add_surface(x = sigma, y = mu)
```


## MLE

We compute the MLE's $(\widehat{\mu}_{\text{MLE}}, \widehat{\sigma}_{\text{MLE}})$ for $(\mu, \sigma)$:

```{r}
xbar <- sum(x)/n
sigma_hat <- sqrt((1 / n) * sum((x - xbar) ^ 2))
c(xbar, sigma_hat)
```


## Interpretation

The MLE $(\widehat{\mu}_{\text{MLE}}, \widehat{\sigma}_{\text{MLE}})$ are the parameter values that best explain/fit the observed data `x`.


# 

## Warm-up

A group of students notes the production code numbers of each of their iphones as follows:

```{r echo = FALSE}
set.seed(301)
sample(1:60000, 6)
```

Using common-sense, use this data to come up with an estimate of the total number of phones produced by Apple in this production run.


## Visualizing Likelihood

Two paradigms for visualizing a function in a computational framework:

1. Connect-the-dots
2. Direct functional representation

Both require writing **functions**.


## Functions in R {.build}

```{r}
my_exponent <- function(a, b) {
  a^b
}
my_exponent(3, 2)
```

- A function is created by the `function()` function.
- Arguments/input go in the parens.
- Code run by function goes in braces.
- Output goes on the final line.
- Run the function to make it available for use.


## Ex. Bernoulli Likelihood {.build}

$$
L(p) = p^{\sum_{i=1}^n x_i}(1 - p)^{n - \sum_{i=1}^n x_i}
$$

```{r}
L_binom <- function(p, x) {
  n <- length(x)
  n_success <- sum(x)
  p ^ n_success * (1 - p) ^ (n - n_success)
}
x <- c(1, 1, 0, 1, 0)
L_binom(p = .5, x = x)
L_binom(3/5, x = x)
```


## Connect-the-dots {.build}

In this approach, you 

1. create a discrete vector of parameter values,
2. run it through the function to get an output vector of the same length, and
3. create a plot by connecting the $(x, y)$ pairs.

```{r}
p_vec <- seq(from = 0, to = 1, by = .005)
head(p_vec)
L_vec <- L_binom(p_vec, x)
head(L_vec)
```


## {.smaller}

```{r}
df <- data.frame(x = p_vec,
                 y = L_vec)
ggplot(df, aes(x = x, y = y)) +
  geom_line()
```

##


## The impact of n

It's helpful to see how the likelihood responds to increasing data.

```{r}
x_big <- rep(x, 10)
L_vec <- L_binom(p_vec, x_big)
df <- data.frame(x = p_vec,
                 y = L_vec)
```


## {.smaller}

```{r}
ggplot(df, aes(x = x, y = y)) +
  geom_line()
```


## Direct function {.smaller}

The second method doesn't require the discretization of the function.


```{r}
ggplot(data.frame(x = c(0, 1)), aes(x)) +
  stat_function(fun = L_binom, args = list(x = x_big))
```

## Back to the Normal {.build .smaller}

$$
l(\mu, \sigma) = -\frac{n}{2}\log(2\pi) - n \log (\sigma) - \frac{1}{2 \sigma^2} \sum_{i = 1}^{n} \left(x_i - \mu\right) ^2
$$

```{r}
l_norm <- function(mu, sigma, x) {
  n <- length(x)
  p1 <- -n / 2 * log(2 * pi)
  p2 <- -n * log(sigma)
  p3 <- -1 / (2 * sigma^2) * sum((x - mu)^2)
  p1 + p2 + p3
}
l_norm_quick <- function(mu, sigma, x) {
  sum(dnorm(x, mean = mu, sd = sigma, log = TRUE))
}
l_norm(0, 5, x)
l_norm_quick(0, 5, x)
```


## Generate data

```{r}
x <- rnorm(100, mean = 0, sd = 5)
```


## Connect the dots

In 2D, we have to define a grid of possible parameter values and compute the likelihood for each coordinate

```{r}
# Define 2D domain
mu <- seq(-1.5, 1.5, length = 500)
sigma <- seq(4, 6, length = 500)
```

```{r}
# Compute log-likelihood
l_surface <- matrix(0, nrow = length(mu), ncol = length(sigma))
for(i in 1:nrow(l_surface)) {
  for(j in 1:ncol(l_surface)) {
    l_surface[i, j] <- l_norm(mu = mu[i], sigma = sigma[j], x)
  }
}
```


##

```{r}
library(plotly)
plot_ly(z = ~l_surface) %>% 
  add_surface(x = sigma, y = mu)
```


## MLE

We compute the MLE's $(\widehat{\mu}_{\text{MLE}}, \widehat{\sigma}_{\text{MLE}})$ for $(\mu, \sigma)$:

```{r}
n <- length(x)
xbar <- sum(x)/n
sigma_hat <- sqrt((1 / n) * sum((x - xbar) ^ 2))
c(xbar, sigma_hat)
```


##

```{r}
set.seed(301)
x <- sample(1:20, 3)
L_unif <- function(beta, x) {
  n <- length(x)
  (1 / beta) ^ n
}
```


## 

```{r}
ggplot(data.frame(x = c(0, 20)), aes(x)) +
  stat_function(fun = L_unif, args = list(x = x))
```

## 

```{r}
ggplot(data.frame(x = c(0, 20)), aes(x)) +
  stat_function(fun = L_unif, args = list(x = x))+
  xlim(c(max(x), 20))
```


