---
title: "MATH 392 Problem Set 8"
output: 
  github_document:
    fig_height: 3
    fig_width: 6
---


```{r, include=FALSE}
require(tidyverse) # load package
require(resampledata)
options(digits=4) # print 4 decimal places only
set.seed(1999) # set seed for reproducibility
```

The core of this assignment is the creation of a MLR simulator in order to investigate the properties of the method. The model under investigation is the following.

$$
Y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \epsilon ; \quad \epsilon \sim N(0, \sigma^2)
$$

You're welcome to select any values that you like for the parameters and any distribution that you like for the $x$'s. The core bit of code should look something like

```{r eval = FALSE}
# set params
B0 <- ___
B1 <- ___
sigma <- ___

# complete specification
n <- ___
epsilon <- r____()
x <- ___

# simulate ys
```


### Sampling distribution

Use your simulator to create an MC approximation of the true sampling distribution of the estimates of $\beta_1$, $E(Y_s)$, and $Y_s$. How do these distribution compare to the analytical form in terms of center, shape, and spread?

### A different model

Consider two variations on the model:

1. An alternate distribution of the $x$'s of your choosing.
2. An alternate distribution for $\epsilon$ (though it still should be centered at 0).

Does the inference change for the statistics investigated in the previous exercise?

### Variance/Covariance

Generate a scatterplot matrix involving all pairwise relationships between the three simulated regression coefficients in the original model. To be clear, this involves generating a good number fitted betas from your simulator and plotting them. Based on a visual assessment of these plots, please characterize the joint distribution of these statistics in terms of center, shape, spread, and covariance.

**Challenge problem**: If you recognize the family of this joint distribution, use your simulated fitted betas to calculate the MLEs of the parameters.

