---
title: "MATH 392 Problem Set 8"
output: 
  github_document:
    fig_height: 3
    fig_width: 6
---


```{r, include=FALSE}
require(tidyverse) # load package
require(resampledata)
options(digits=4) # print 4 decimal places only
set.seed(1999) # set seed for reproducibility
```

The core of this assignment is the creation of a MLR simulator in order to investigate the properties of the method. The model under investigation is the following.

$$
Y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \epsilon ; \quad \epsilon \sim N(0, \sigma^2)
$$

You're welcome to select any values that you like for the parameters and any distribution that you like for the $x$'s. The core bit of code should look something like

```{r eval = FALSE}
# set params
B0 <- ___
B1 <- ___
B2 <- ___
sigma <- ___

# complete specification
n <- ___
x <- ___

# simulate ys (this part inside a for loop)
epsilon <- r____()
```


### Part I. Sampling distributions

Use your simulator to create an MC approximation of the true sampling distribution of the estimates of $\beta_1$, $E(Y_s)$, and $Y_s$ corresponding to a fixed new observation $x_s$. How do these distribution compare to the analytical form in terms of center, shape, and spread?

### Part II. A different model

Consider two variations on the model:

1. An alternate distribution of the $x$'s of your choosing.
2. An alternate distribution for $\epsilon$ (though it still should be centered at 0).

Does the inference change for the statistics investigated in the previous exercise? For this part, you should be able to recycle much of your code from the previous part.

### Part III. Variance/Covariance

Generate a scatterplot matrix involving all pairwise relationships between the three simulated regression coefficients in the original model. To be clear, this involves generating a good number fitted betas from your simulator and plotting them. Based on a visual assessment of these plots, please characterize the joint distribution of these statistics in terms of center, shape, spread, and covariance.

