---
title: "MATH 392 Problem Set ZZQ"
author: "REPLACE WITH YOUR NAME"
date: "REPLACE WITH TODAY'S DATE"
output: 
  pdf_document:
    fig_height: 3
    fig_width: 6
---


```{r, include=FALSE}
require(tidyverse) # load packages
require(resampledata)
options(digits=4) # print 4 decimal places only
set.seed(1999) # set seed for reproducibility
```

## Problem X

### Analytic solution

Math can be typed in as Latex either inline with single dollar signs $\bar{x} = \frac{1}{n} \sum_{i=1}^n x_i$ or offset with double dollar signs.

$$
H_0: \rho = 0
$$

or if you'd like to align multiple lines by the equal sign:

$$
\begin{aligned}
A &= x \\
B &= y + 7
\end{aligned}
$$

A very useful resource if you're just staring with \LaTeX is \texttt{http://detexify.kirelabs.org/classify.html}.

### Empirical solution

```{r}
# code goes here
```

## Problem Y

### Analytic solution

### Empirical solution

```{r}
# code goes here
```


## Problem Z

### Analytic solution

### Empirical solution

```{r}
# code goes here
```

